### 2016-11-22 Projects/DelCo_2012/Report/DelCo_2012.tex
edits to CAF report from JH comments

\section{Description of Experiments}
\label{sec:desc_experiments}
The experimental setup for this test series is described by the three types of experiments conducted: spray density measures, gas layer cooling, and fire suppression.  Baseline operating conditions were agreed upon at the CAFS workshop for these experiments (Table~\ref{tab:op_condition})~\cite{Grant:2011}. The nozzles chosen for these experiments were Task Force Tips (TFT) Metro 1, fixed gallonage nozzles with a flow rate of 120~gpm. These are automatic nozzles which have a 22~mm (0.88~in) slug-tip and can be varied from a straight stream to a wide fog stream for application. The setup also included 30.5~m (100~ft) of 44~mm (1.75~in) diameter hose. Nozzles were mounted to a ``blitz-fire'' monitor device and placed inside the structures. The nozzles were controlled and operated via a gate valve from the exterior. The exterior control had a gated wye to allow the flow of CAF through a hose directed away from the test structure. When the flow of CAF was needed for the experiment, the gated wye would be switched to allow the CAF to flow into the test structure. This was done to minimize the build-up of ``dry'' foam or a slug upon the initial flow of the nozzle into the test structure.

@@

In Figure~\ref{fig:fs_hall_a1}, the red dashed line (linear fit to the data, see Section~\ref{sec:Gas_Cooling}) closely follows the black line, which indicates near equal cooling between CAF and water. The R$^2$ value of 0.901 shows a strong fit to the data. When the second nozzle (door nozzle) was activated, the temperature differences are noticeably lower compared to the hallway nozzle, despite it being an indirect attack. While the fit is not as strong (R$^2$~=~0.643), the red line still falls within the experimental uncertainty, meaning that differences in the impact of CAF and water are indistinguishable. Despite being closer to the nozzles, thermocouple tree A2 shows similar temperature difference results from the hallway nozzle activation to the A1 array (see Figure~\ref{fig:fs_door_a1}). At array A2 for the door nozzle (see Figure~\ref{fig:fs_door_a2}), the red dashed falls outside of the experimental uncertainty, indicating a slight advantage of cooling for CAF.

@@

There were seven suppression tests conducted in the two-story, purpose-built structure. All seven of the tests used a fuel load consisting of three sofas, polypropylene carpeting over polyurethane foam padding, and OSB paneling along three walls and the ceiling. Five of the seven tests also had two stacks of wooden pallets in addition to the initial fuel load (Section~\ref{sec:fire_suppression_furniture_fuel}). Ignition occurred in the southwest corner of the ground floor and available ventilation was through the ground floor double doors and the door on the south side of the second story of the structure. The fires were intended to replicate room and contents basement fires typically seen within residential structures. Suppression occurred at the basement double doors of the structure. Table~\ref{tab:Test_Descriptions_2} provides an overview of the suppression tests conducted.

\begin{table}[!ht]
\centering
\footnotesize
\caption{Two-Story Fire Suppression Test Descriptions}\label{tab:Test_Descriptions_2}
\begin{tabular}{clllll}
\toprule[1.5pt]
Test $\#$  & Name	& Stream			& Flow Rate		& Fuel                    & Ventilation  \\
\midrule
 13  & CAF~7     &  Straight Stream  	&  120~gpm/60~cfm   & Furniture           & Basement Doors, 1st Floor Door \\
 14  & Water~7   &  Straight Stream  	&  120~gpm    		& Furniture           & Basement Doors, 1st Floor Door \\
 15  & CAF~8a    &  Straight Stream  	&  120~gpm/60~cfm   & Furniture + Pallet  & Basement Doors, 1st Floor Door \\
 16  & CAF~8b    &  Straight Stream  	&  120~gpm/60~cfm   & Furniture + Pallet  & Basement Doors, 1st Floor Door \\
 17  & Water~8   &  Straight Stream     &  120~gpm          & Furniture + Pallet  & Basement Doors, 1st Floor Door \\
 18  & Class A~1 &  Straight Stream  	&  120~gpm		    & Furniture + Pallet  & Basement Doors, 1st Floor Door \\
 19  & CAF~9a    &  1 3/8'' Ball    	&  120~gpm/60~cfm   & Furniture + Pallet  & Basement Doors, 1st Floor Door \\
 20  & CAF~9b    &  Solid Stream        &  120~gpm/60~cfm   & Furniture + Pallet  & Basement Doors, 1st Floor Door \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

The time history profile of the temperature response along with the heat flux at three positions is shown in Figures~\ref{fig:water7_tca10}--\ref{fig:caf9a_hf}. Figures~\ref{fig:water7_tca10} and \ref{fig:water7_hf} show the results for test Water~7, while Figures~\ref{fig:caf9a_tca10} and \ref{fig:caf9a_hf} show the results for test CAF~9a. The remainder of the fire suppression test data and results are included in Appendix~\ref{app:fire_suppression2}.

@@

While CAF has been readily accepted for use in the wildland and wildland urban interface, the effectiveness of CAF for interior structural fire fighting varies widely. Questions remain regarding whether CAF provides improved gas cooling when compared to plain water. During the planning of the experiments for this project, there were a range of hypotheses regarding potential outcomes. Some technical panel members thought that the excellent heat absorbing capacity of plain water would exceed the heat absorbing capabilities of CAF. Their position was based on the fact that CAF has injected air. Because air is a great insulator, it slows heat transfer and results in less cooling. It was their belief that the reduced gas cooling capabilities of CAF may have led to firefighter injuries during interior fire attacks. Other members of the technical panel thought that the bubble structure would provide the same mass flow rate of water but with more surface area. The increased surface area would enable faster heat transfer rate and therefore improve the cooling of the hot gas layer. There were also thoughts that the CAFS would coat the ceiling and eliminate energy feedback from the ceiling to the hot gas layer, thereby providing enhanced cooling.

The gas cooling experiments were designed to measure the ability of water streams and CAF streams to cool a hot gas layer without any application of agent to the source of the heat. The 3.35~m (11~ft) tall ceiling in the burn building allowed a hot gas layer of at least 1.83~m (6 ft) to form. During the testing, a minimum temperature of 250~$^{\circ}$C (482~$^{\circ}$F) at 1.83~m (6~ft) below the ceiling was attained prior to the beginning of gas cooling. In each gas cooling experiment, the same mass flow rate of water was used for both the plain water stream and the CAF stream. Typically, the volumetric flow rate of water was 7.57~L/s (120~gpm). The density of water is 1000~kg/m$^3$ (8.3~lb/gal), therefore the mass flow rate was 7.57~kg/s (16.7~lb/s). The average time of flow was approximately 15~s, therefore a total of 114~kg (251~lb) of water was introduced into each gas cooling experiment.

@@

The estimated total expanded uncertainty of the temperature measurements in these experiments is $\pm$~15~\% (see Section~\ref{subsec:Uncertainty}). So, for the difference between two temperature measurements to be considered significantly different, the difference would have to be greater by more than $\pm$~15~\%. If the difference between two measurements was $\pm$~15~\% or less, the results are considered similar. That said, the estimated uncertainty is based on previous studies where suppression with a water based agent was included.

As mentioned in the results section, the conditions at TC Array 3 represented the ``best'' gas cooling results because due to the array's position, it was impacted the least by agents hitting and cooling the thermocouples themselves. Under the given conditions of a residential scale room and the amount of water introduced into the hot gas layer, it should be no surprise that the water is absorbing the energy from the upper layer. Air filled bubbles in the CAF stream had no discernible impact, either positive or negative, in the gas cooling experiments. At the same time, there was no measurable negative effect of the air bubbles with regard to gas cooling.

@@

\subsection{Iowa State University Fire Flow Formula (1963)}
The Iowa Formula for determining the required fire flow rate for a given structure was developed by Keith Royer and Bill Nelson and is based on theoretical heat absorption capabilities of water. The required fire flow rate ($\dot{V}$) in gpm is equal to the volume ($V$) of the fire compartment in cubic feet divided by 100:

@@

Due to inefficiencies in the application of water, some feel that the fire flow value from the Iowa State University formula should be multiplied by a factor of two to four~\cite{NFPA}. Applying the formula based on floor area yields an application density of 3.3~l/min/m$^2$ (0.08~gpm/ft$^2$) for ceiling heights of 2.44~m (8~ft) or 4.1~l/min/m$^2$ (0.10~gpm/ft$^2$) for ceiling heights of 3.05~m (10~ft). The fire room compartment volumes in these experiments ranged from approximately 66~m$^3$ (2330~ft$^3$) to 125~m$^3$ (4410~ft$^3$). Therefore, the required flow rate for the fire room in the single-story structure based on the Iowa State University formula was approximately 87~l/min (23~gpm), and the required flow rate for the fire room in the lower level of the two-story structure was approximately 168~l/min (44~gpm). During the experiments, the flow rate was approximately 454~l/min (120~gpm), which exceeds the required flow rate of the Iowa State Formula. The fact that all of the fires were quickly controlled during the experiments is consistent with this methodology.

\subsection{National Fire Academy (1981)}
The National Fire Academy (NFA) developed a formula based on a study of fire flows that have been successful in controlling a large number of working fires~\cite{Klaene:1}. This formula sets the required fire flow ($\dot{V}$) in gpm equal to the floor area ($A$) in square feet divided by 3.

@@

Both of the formulas listed above do not explicitly account for the ventilation of the fire structure, only the size of the fire structure. Further, the most recent of the two formulas was developed in 1981, based on the fire ground experience of fire chiefs. This experience was likely from fires that occurred in the 1960's and the 1970's. It has been demonstrated that the synthetic fuels which fill the homes of the 2000's burn faster and with higher heat release rates than the natural fuels that filled the homes in the 1940's and 1950's. The change of materials in homes began in the 1960's and has continued to present day~\cite{Kerber:2}. It was unlikely, however, that this change was factored into the NFA formula.

@@

All of the fire conditions generated during these experiments were ventilation limited or ventilation controlled fires. Essentially, the conditions in the compartment were fuel rich. If the ventilation was increased to allow additional oxygen to be supplied to the fuel gases in the fire compartment, the heat release rate of the fire would have increased. There are simple formulas which may provide some insight here, although this simple algorithm has not been validated with the geometries that were used in the experiments described in this report. The purpose here is to get a sense of the amount of heat energy that the hose streams needed to overcome. This algorithm, like the NFA formula, has never been validated for this application.

@@

In all the suppression experiments, the fire was controlled with a total flow of 227~L (60~gal) or less of agent. During the course of the suppression experiments, the fuel packages were modified to incorporate more wood-based fuels. During the planning of the experiments, some of the technical panel members felt that synthetic fuels, such as polyurethane foam, burned mostly at the surface and were therefore easily cooled and suppressed. Wood-based fuels, however, develop a char layer. Therefore the burning of wood-based fuels would be ``deep seated'', which might provide a more significant suppression challenge. The reduced surface tension of Class A foam solution could allow the solution to better penetrate and cool the burning wood-based fuels. It was thought that with more exposed burning wood surfaces in the fuel load, CAF would have a better the opportunity to demonstrate a difference in suppression. As shown in Chapter~\ref{sec:fuel_load}, the wood fuel load consisted of stacks of wood pallets, plywood lined walls, and an exposed wood ``ceiling'' that was constructed of wood ``i-beams'' and OSB. The experiments conducted with the wood fuel load did not demonstrate a significant difference between water and CAF.

The final set of suppression experiments used a larger space than the previous two sets of suppression experiments in the single-story structure. As discussed previously, with the larger area and double the ventilation of the previous experiments, the fire should be more of a challenge to the hose streams. There was also another variable that would be introduced: a firefighter controlling the nozzle. Recall that in the single-story experiments, a monitor nozzle was used in a fixed position. As a result, after the initial fire knockdown, it was typical that the agent was not able to hit all surface of the burning fuel sufficiently and a rekindle would occur. This can be seen for both the water and CAF experiments in the data provided in Appendix~\ref{app:fire_suppression}. In some single-story experiments, the fire was given time to redevelop and additional nozzle flows from the fixed positions were conducted to collect more data and to see if the fires could be extinguished. During the additional flows, no significant differences were seen. In all of the single-story experiments firefighters with a hoseline were needed to extinguish any remaining fire and hot spots. For the last set of experiments, the choice was made to use a firefighter so that the nozzle would be moved in an informed manner in order to make the best use of the hose stream to suppress the fire in the lower level of the two-story structure.

@@

During the workshop conducted for this project~\cite{Grant:2011}, it was suggested that CAF could suppress, or knockdown, a fire three to five times faster than water. The data collected during this study does not demonstrate significant differences between CAF and water for use in interior structural fire fighting on a residential scale. Going back to the premise of the study, preventing firefighters from being injured during interior fire fighting operations with CAF, there is nothing in these fire suppression results to suggest that CAF would present a thermal hazard, provided that an adequate amount of water is being flowed for the given fire condition.

@@

Eighty-eight gas cooling experiments were conducted to study the effectiveness of cooling the upper gas layer in a space adjacent to the area of fire involvement. These tests examined five different hose stream patterns, two nozzle types, and both compressed air foam and water as the suppression agent. The focus of this experimental series was to study the impact of the test variables on the hot gas layer temperatures when a fixed hose stream was applied to a room adjacent to the fire room. Suppression agents were not applied to the fire in the burn room. In this report, an aggregate analysis was made to examine differences due to the location of the instrumentation relative to the impact area of the agent and compare the differences in the amount of gas cooling between plain water and CAF hose streams. No significant differences were found in the gas cooling capabilities of the two agents. A more detailed analysis of the individual experiments was conducted by Mitchell of Cal Poly~\cite{Mitchell:1}. The conclusion of both analyses is in agreement. The full gas cooling data set can be found in the Mitchell report.

@@

The Delaware County Emergency Services Training Center provided the location and logistical support for these experiments. The Delaware County team was under the direction of Kerby Kerber.  For construction and test support the team included: Joe Bingham, Dave Wright, Dave Nercesian (small engine repair), Mike Gura, Scott Wiercinski, Bill Fields (Fields Fire Protection), John Hale and Todd John.

We also thank the dedicated team of fire instructors and firefighters that supported these experiments; Brian Righter (Rudy), John McGowan, Matthew Poissant, John Frey, Brian Embert, Raymond Keller III, Mike Bramble, Edward McBride, Mark Morrissey, Bill Norris, and Matt LeTourneau.

>>>>

\section{Description of Experiments}
\label{sec:desc_experiments}
The experimental setup for this test series is described by the three types of experiments conducted: spray density measures, gas layer cooling, and fire suppression.  Baseline operating conditions were agreed upon at the CAFS workshop for these experiments (Table~\ref{tab:op_condition})~\cite{Grant:2011}. The nozzles chosen for these experiments were Task Force Tips (TFT) Metro 1 \textregistered, fixed gallonage nozzles with a flow rate of 120~gpm. These are automatic nozzles which have a 22~mm (0.88~in) slug-tip and can be varied from a straight stream to a wide fog stream for application. The setup also included 30.5~m (100~ft) of 44~mm (1.75~in) diameter hose. Nozzles were mounted to a Blitz-Fire \textregistered monitor device and placed inside the structures. The nozzles were controlled and operated via a gate valve from the exterior. The exterior control had a gated wye to allow the flow of CAF through a hose directed away from the test structure. When the flow of CAF was needed for the experiment, the gated wye would be switched to allow the CAF to flow into the test structure. This was done to minimize the build-up of ``dry'' foam or a slug upon the initial flow of the nozzle into the test structure.

@@

In Figure~\ref{fig:fs_hall_a1}, the red dashed line (linear fit to the data, see Section~\ref{sec:Gas_Cooling}) closely follows the black line, which indicates near equal cooling between CAF and water. The R$^2$ value of 0.879 shows a strong fit to the data. When the second nozzle (door nozzle) was activated, the temperature differences are noticeably lower compared to the hallway nozzle (indirect attack). While the fit is not as strong (R$^2$~=~0.643), the red line still falls within the experimental uncertainty, meaning that differences in the impact of CAF and water are indistinguishable. Despite being closer to the nozzles, thermocouple tree A2 shows similar temperature difference results from the hallway nozzle activation to the A1 array (see Figure~\ref{fig:fs_door_a2}). At array A2 for the door nozzle (see Figure~\ref{fig:fs_door_a2}), the red dashed falls outside of the experimental uncertainty, indicating a slight advantage of cooling for CAF.

@@

There were seven suppression tests conducted in the two-story, purpose-built structure. All seven of the tests used a fuel load consisting of three sofas, polypropylene carpeting over polyurethane foam padding, and OSB paneling along three walls and the ceiling. Five of the seven tests also had two stacks of wooden pallets in addition to the initial fuel load (Section~\ref{sec:fire_suppression_furniture_fuel}). Ignition occurred in the southwest corner of the ground floor and available ventilation was through the ground floor double doors and the door on the south side of the second story of the structure. The fires were intended to replicate room and contents basement fires typically seen within residential structures. Suppression occurred at the basement double doors of the structure. Table~\ref{tab:Test_Descriptions_2} provides an overview of the suppression tests conducted. In all cases, ventilation was provided through open basement doors and a door on the first floor.

\begin{table}[!ht]
\centering
\small
\caption{Two-Story Fire Suppression Test Descriptions}\label{tab:Test_Descriptions_2}
\begin{tabular}{cllll}
\toprule[1.5pt]
Test $\#$  & Name	& Stream			& Flow Rate		& Fuel                    \\
\midrule
 13  & CAF~7     &  Straight Stream  	&  120~gpm/60~cfm   & Furniture           \\
 14  & Water~7   &  Straight Stream  	&  120~gpm    		& Furniture           \\
 15  & CAF~8a    &  Straight Stream  	&  120~gpm/60~cfm   & Furniture + Pallet  \\
 16  & CAF~8b    &  Straight Stream  	&  120~gpm/60~cfm   & Furniture + Pallet  \\
 17  & Water~8   &  Straight Stream     &  120~gpm          & Furniture + Pallet  \\
 18  & Class A~1 &  Straight Stream  	&  120~gpm		    & Furniture + Pallet  \\
 19  & CAF~9a    &  1 3/8'' Ball    	&  120~gpm/60~cfm   & Furniture + Pallet  \\
 20  & CAF~9b    &  Solid Stream        &  120~gpm/60~cfm   & Furniture + Pallet  \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

The time history profile of the temperature response along with the total heat flux at three positions is shown in Figures~\ref{fig:water7_tca10}--\ref{fig:caf9a_hf}. Figures~\ref{fig:water7_tca10} and \ref{fig:water7_hf} show the results for test Water~7, while Figures~\ref{fig:caf9a_tca10} and \ref{fig:caf9a_hf} show the results for test CAF~9a. The remainder of the fire suppression test data and results are included in Appendix~\ref{app:fire_suppression2}.

@@

While CAF has been readily accepted for use in the wildland and wildland urban interface, the effectiveness or advantages of CAF for interior structural fire fighting was undocumented by experimental data. Questions remain regarding whether CAF provides improved gas cooling when compared to plain water. During the planning of the experiments for this project, there were a range of hypotheses regarding potential outcomes. Some technical panel members thought that the excellent heat absorbing capacity of plain water would exceed the heat absorbing capabilities of CAF. Their position was based on the fact that CAF has injected air. Because air is a great insulator, it slows heat transfer and results in less cooling. It was their belief that the reduced gas cooling capabilities of CAF may have led to firefighter injuries during interior fire attacks. Other members of the technical panel thought that the bubble structure would provide the same mass flow rate of water but with more surface area. The increased surface area would enable faster heat transfer rate and therefore improve the cooling of the hot gas layer. There were also thoughts that the CAFS would coat the ceiling and eliminate energy feedback from the ceiling to the hot gas layer, thereby providing enhanced cooling. However, CAF has not been widely used for structure fires.

The gas cooling experiments were designed to measure the ability of water streams and CAF streams to cool a hot gas layer without any application of agent to the source of the heat. The 3.35~m (11~ft) tall ceiling in the burn building allowed a hot gas layer of at least 1.83~m (6 ft) to form. During the testing, a minimum temperature of 250~$^{\circ}$C (482~$^{\circ}$F) at 1.83~m (6~ft) below the ceiling was attained prior to the beginning of gas cooling. In each gas cooling experiment, the same mass flow rate of water was used for both the plain water stream and the CAF stream. Typically, the volumetric flow rate of water was 7.57~L/s (120~gpm). The density of water is 1000~kg/m$^3$ (8.3~lb/gal), therefore the mass flow rate was 7.57~kg/s (16.7~lb/s). The average time of flow was approximately 15~s. A total of 114~kg (251~lb) of water was introduced into each gas cooling experiment.

@@

The estimated total expanded uncertainty of the temperature measurements in these experiments is $\pm$~15~\% (see Section~\ref{subsec:Uncertainty}). For the difference between two temperature measurements to be considered significantly different, the difference would have to be greater by more than $\pm$~15~\%. If the difference between two measurements was $\pm$~15~\% or less, the results are considered similar. The estimated uncertainty is based on previous studies where suppression with a water based agent was included.

As mentioned in the results section, the conditions at TC Array 3 represented the ``best'' gas cooling results. It was impacted the least by agents hitting and cooling the thermocouples themselves. Under the given conditions of a residential scale room and the amount of water introduced into the hot gas layer, it should be no surprise that the water is absorbing the energy from the upper layer. Air filled bubbles in the CAF stream had no discernible impact, either positive or negative, in the gas cooling experiments. At the same time, there was no measurable negative effect of the air bubbles with regard to gas cooling.

@@

\subsection{Iowa State University Fire Flow Formula}
The Iowa Formula for determining the required fire flow rate for a given structure was developed by Keith Royer and Bill Nelson and is based on theoretical heat absorption capabilities of water \cite{Royer:ISU}. The required fire flow rate ($\dot{V}$) in gpm is equal to the volume ($V$) of the fire compartment in cubic feet divided by 100:

@@

Due to inefficiencies in the application of water, some feel that the fire flow value from the Iowa State University formula should be multiplied by a factor of two to four~\cite{NFPA}. Applying the formula based on floor area yields an application density of 3.3~l/min/m$^2$ (0.08~gpm/ft$^2$) for ceiling heights of 2.44~m (8~ft) or 4.1~l/min/m$^2$ (0.10~gpm/ft$^2$) for ceiling heights of 3.05~m (10~ft). The fire room compartment volumes in these experiments ranged from approximately 66~m$^3$ (2330~ft$^3$) to 125~m$^3$ (4410~ft$^3$). The required flow rate for the fire room in the single-story structure based on the Iowa State University formula was approximately 87~l/min (23~gpm), and the required flow rate for the fire room in the lower level of the two-story structure was approximately 168~l/min (44~gpm). During the experiments, the flow rate was approximately 454~l/min (120~gpm), which exceeds the required flow rate of the Iowa State Formula. The fact that all of the fires were quickly controlled during the experiments is consistent with this methodology.

\subsection{National Fire Academy}
The National Fire Academy (NFA) developed a formula based on a study of fire flows that have been successful in controlling a large number of working fires~\cite{Klaene:1}. This formula sets the required fire flow ($\dot{V}$) in gpm equal to the floor area ($A$) in square feet divided by 3.

@@

Both of the formulas listed above do not explicitly account for the ventilation of the fire structure, only the size of the fire structure. The most recent of the two formulas was developed in 1981, based on the fire ground experience of fire chiefs. This experience was likely from fires that occurred in the 1960's and the 1970's. It has been demonstrated that the synthetic fuels which fill the homes of the 2000's burn faster and with higher heat release rates than the natural fuels that filled the homes in the 1940's and 1950's. The change of materials in homes began in the 1960's and has continued to present day~\cite{Kerber:2}. It was unlikely, however, that this change was factored into the NFA formula.

@@

All of the fire conditions generated during these experiments were ventilation limited or ventilation controlled fires. Essentially, the conditions in the compartment were fuel rich. If the ventilation was increased to allow additional oxygen to be supplied to the fuel gases in the fire compartment, the heat release rate of the fire would have increased. There are simple formulas which may provide some insight here, although this simple algorithm has not been validated with the geometries that were used in the experiments described in this report. The purpose here is to estimate the amount of heat energy that the hose streams needed to overcome. This algorithm, like the NFA formula, has never been validated for this application. The heat release rate is:

@@

In all the suppression experiments, the fire was controlled with a total flow of 227~L (60~gal) or less of agent. During the course of the suppression experiments, the fuel packages were modified to incorporate more wood-based fuels. During the planning of the experiments, some of the technical panel members felt that synthetic fuels, such as polyurethane foam, burned mostly at the surface and were therefore easily cooled and suppressed. Wood-based fuels, however, develop a char layer. The burning of wood-based fuels would be ``deep seated'', which might provide a more significant suppression challenge. The reduced surface tension of Class A foam solution could allow the solution to better penetrate and cool the burning wood-based fuels. It was thought that with more exposed burning wood surfaces in the fuel load, CAF would have a better the opportunity to demonstrate a difference in suppression. As shown in Section~\ref{sec:fuel_load}, the wood fuel load consisted of stacks of wood pallets, plywood lined walls, and an exposed wood ``ceiling'' that was constructed of wood ``i-beams'' and OSB. The experiments conducted with the wood fuel load did not demonstrate a significant difference between water and CAF.

The final set of suppression experiments used a larger space than the previous two sets of suppression experiments in the single-story structure. As discussed previously, with the larger area and double the ventilation of the previous experiments, the fire should have been more challenging to the hose streams. There was also another variable that was be introduced: a firefighter controlling the nozzle. Recall that in the single-story experiments, a monitor nozzle was used in a fixed position. As a result, after the initial fire knockdown, it was typical that the agent was not able to hit all surface of the burning fuel sufficiently and a rekindle would occur. This can be seen for both the water and CAF experiments in the data provided in Appendix~\ref{app:fire_suppression}. In some single-story experiments, the fire was given time to redevelop and additional nozzle flows from the fixed positions were conducted to collect more data and to see if the fires could be extinguished. During the additional flows, no significant differences were seen. In all of the single-story experiments firefighters with a hoseline were needed to extinguish any remaining fire and hot spots. For the last set of experiments, the choice was made to use a firefighter so that the nozzle would be moved in an informed manner in order to make the best use of the hose stream to suppress the fire in the lower level of the two-story structure.

@@

During the workshop conducted for this project~\cite{Grant:2011}, it was suggested that CAF could suppress, or knockdown, a fire three to five times faster than water. The data collected during this study does not demonstrate significant differences between CAF and water for use in interior structural fire fighting on a residential scale. There is nothing in these fire suppression results to suggest that CAF would present a thermal hazard, provided that an adequate amount of water is being flowed for the given fire condition.

@@

Eighty-eight gas cooling experiments were conducted to study the effectiveness of cooling the upper gas layer in a space adjacent to the area of fire involvement. These tests examined five different hose stream patterns, two nozzle types, and both compressed air foam and water as the suppression agent. The focus of this experimental series was to study the impact of the test variables on the hot gas layer temperatures when a fixed hose stream was applied to a room adjacent to the fire room. Suppression agents were not applied to the fire in the burn room. In this report, an aggregate analysis was made to examine differences due to the location of the instrumentation relative to the impact area of the agent and compare the differences in the amount of gas cooling between plain water and CAF hose streams. The results showed that water was generally more effective than CAF, though both agents were successful with respect to gas cooling and fire suppression. A more detailed analysis of the individual experiments was conducted by Mitchell of Cal Poly~\cite{Mitchell:1}. The conclusion of both analyses is in agreement. The full gas cooling data set can be found in the Mitchell report.

@@

The Delaware County Emergency Services Training Center provided the location and logistical support for these experiments. The Delaware County team was under the direction of Kerby Kerber.  The construction and test support team included: Joe Bingham, Dave Wright, Dave Nercesian (small engine repair), Mike Gura, Scott Wiercinski, Bill Fields (Fields Fire Protection), John Hale and Todd John.

Thanks are given to the dedicated team of fire instructors and firefighters that supported these experiments; Brian Righter (Rudy), John McGowan, Matthew Poissant, John Frey, Brian Embert, Raymond Keller III, Mike Bramble, Edward McBride, Mark Morrissey, Bill Norris, and Matt LeTourneau.
